Learning Actionable Analytics (with applications for reducing defects and reducing runtimes)

### 2015-05-02 manuscript.tex
Updates from ShareLaTeX

\author{
% You can go ahead and credit any number of authors here,
% e.g. one 'row of three' or two rows (consisting of one row of three
% and a second row of one, two or three).
%
% The command \alignauthor (no curly braces needed) should
% precede each author name, affiliation/snail-mail address and
% e-mail address. Additionally, tag each line of
% affiliation/address with \affaddr, and tag the
% e-mail address with \email.
%
% 1st. author
Rahul Krishna, Tim Menzies, Xipeng Shen\\
       \affaddr{Computer Science, NcState, USA}\\
       \{i.m.ralk, tim.menzies, xipengshen\}@gmail.com
% 3rd. author
 % use '\and' if you need 'another row' of author names
% 4th. author
\and
 Andrian Marcus\\
       \affaddr{Computer Science, UtDallas, USA}\\
       {amarcus@utdallas.edu} }

\begin{document}
  \title{Learning Actionable Analytics 
 (with applications for reducing defects and reducing runtimes)}

  % make the title area
  \maketitle
  \pagestyle{plain}

@@

\section{Introduction}
Business  users are now demanding that data mining tools
be augmented with tools to support  business-level
interpretation of that data. For example,
at a  panel on software analytics at ICSE'12,
industrial practitioners lamented the state of the art in data mining
and software engineering~\cite{menzies12a}. Panelists commented that
``prediction is all well and good, but what about decision
making?''. That is, these panelists are more interested in the interpretations
that follow the mining, rather than just  the mining. For example:
\bi
\item Instead of just accepting  {\em predictions} on how many 
 software defects
to expect,  business users might now demand a {\em plan} to
reduce the likelihood of those defects;
\item Instead of just accepting {\em predictions} on the runtime
time of their software, business users might now demand
a {\em plan} to reduce that runtime.
\ei
In these two points, we say that a ``plan'' is a  set of proposed changes in order to better achieve some goal. 
Generating {\em plans} is a  very different task the standard
data mining tasks of {\em predicting} some discrete or numeric class.
Predictors
and classifiers only tell us what {\em is} while planners 
  tell us {\em what to change}. For the kind of business user
discussed above, planners are the preferred to mere prediction systems
since, as one gruff user once shouted at us:
\begin{quote}
{\em ``Don't tell me what \underline{is};
tell me what to \underline{change}.''}
\end{quote}
In order to satisfy the demands of this new kind of business user, we  study the generation of {\em plans} in the context
of software analytics.  
As reported in this paper, that study makes five contributions:
\be
\item A general framework for augmenting a prediction system such that
it becomes a planning system;
\item A  test that recognizes  when we should {\em not}  plan: in summary, our kind of planning is
not recommended in domains that  generate poorly performing predictors;
\item A general verification rig for assessing the value of plans generated within this rig.
\item A novel near-linear-time instance-based planner called ``HOW''. 
\item An experimental validation that the HOW planner is useful for reducing the
expected values of 
\bi
\item Software defects in certain object-oriented JAVA systems listed in \tion{oo}:  Ant Camel, Ivy, Jedit, Log4, Lucene, Poi, Synapse, Velocity, Xalan;
\item The runtime of the software systems discussed in \tion{cpu}: Berkeley DB, Apache, SQLite, LLVM, 
  x264.
\ei
\ei
HOW is an {\em instance-based planner} in that it make its conclusions by
studying the space of examples. This is a very different approach to
{\em model-based planners} that use the model
as sub-routine within a Monte Carlo~\cite{me07f} or evolutionary 
programming rig~\cite{krall14,harman12dec}. 
We prefer HOW's instance-based approach since
it is not dependent on any underlying model.
This is important since:
\bi
\item
When some models are hard to access or slow to run (see examples, below),
then it useful to reason from historical logs rather than from hard-to-collect
model output.
\item
For reasons of confidentionality or priorpiarty restrictions, 
It can be harder to access 
\item
Building and certifying a model   can take years. For example, in previous work~\cite{me07f} we used models
developed by Boehm's group at the University of Southern California.
Some of those models took decades to develop and mature (from 1981~\cite{boehm81} to 2000~\cite{boehm00b}). 
\ei
The rest of this paepr is structued as follows.

\section{From ``Prediction'' to ``Planning''}

approach if the underlying model

This paper presents the following general framework for extending  predictor systems
such that they become 
policy generators:

\begin{small}
\[
\begin{array}{r}
project\\
data
\end{array}\rightarrow \left\{\begin{array}{l}\mathit{train}
\rightarrow \left\{\begin{array}{l}\mathit{build\;predictor\;(e.g.\;CART\;or\; RF)}\\
\mathit{build\;planner\; (e.g.\; HOW)}\end{array}\right.
\\
~\\
\mathit{test}  \rightarrow  \left\{\begin{array}{l@{~}l}
\mathit{before}& =\mathit{predictor}(\mathit{test})\\
\mathit{if\; before} & \ge  \mathif{satisfactory\; then}\\
 &
\begin{array}{l@{~}l}

\mathit{test'}& = \mathit{planner}(\mathit{test})\\
\mathit{after}& =\mathit{predictor}(\mathit{test'})\\ 
\Delta& =\mathit{after} / \mathit{before}
\end{array}
\end{array}\right.
\end{array} \right.
\]
\end{small}
In the above:
\bi
\item
Project data is divided into a train and test set.
\item
From the train set, we learn {\em both} a policy generator as well
as a predictor using (e.g.)
CART (for making numeric predictions~\cite{Breiman1984}) or Random Forests (for making predictions
about symbolic classes~\cite{Breiman2001}). 
\item
The test data is then studied to generate two predictions:
a {\em before} prediction (on the raw test data); then an {\em after}
prediction (after applying the policies learned from the training data).
\item
The improvement delta $\Delta$ is then computed by dividing {\em after} and {\em before}.
\ei
This paper presents a novel policy generator called ``HOW''.
HOW is a near-linear time  instance-based
local search algorithm that looks for improvement trends in the data,
then adjusting examples to nudge them along that trend.
The experiments of this paper apply HOW to learning policies for 
software defect reduction and for compiling software that runs faster.
At least for the data explored here, HOW's reductions in defects and runtimes
are very large-- up to 50\% fewer defects and up to 50\% faster runtimes.

@@

\subsection{Defect Data Set}
The data was obtained from the PROMISE repository\footnote{Promise Repository: \url{http://openscience.us/repo}}. For the defect data we investigated 32 releases from 11 open source Java projects defined by the metrics highlighted in figure ~\ref{fig:ck}: \kw{Apache Ant} (1.5 -- 1.7), \kw{Apache Camel} (1.2 -- 1.6), \kw{Apache Ivy} (1.1 -- 2.0), \kw{JEdit} (4.1 -- 4.3), \kw{Apache Log4j} (1.0 -- 1.2), \kw{Apache Lucene} (2.0 -- 2.2), \kw{PBeans} (1.0 and 2.0), \kw{Apache POI} (2.0 -- 3.0), \kw{Apache Synapse} (1.0 -- 1.2), \kw{Apache Velocity} (1.4 -- 1.6), and \kw{Apache Xalan-Java} (2.5 -- 2.7).

@@

{\bfseries RQ3} asks if the aforementioned method can be expanded to other kinds of data. To answer this, we have used the performance prediction data set collected by Siegmund et al. for their ICSE '12 paper~\cite{sven12}. The data sets have been deployed as a part of the \textsc{SPLConqueror} tool\footnote{\url{http://fosd.de/SPLConqueror}}. This data set contains six examples of real world configurable systems that cover a broad spectrum of scenarios, see figure \ref{fig:cpm} for a summary. The systems have different characteristics (lines of code: 40 thousand to 300 thousand; Valid configurations: 192 to about 3.9 million), different implementation languages (C, C++, and Java), different mechanism (conditional compilation, configuration files, and command-line options), and different domains (Web servers, compilers, encoders, and databases). The data contains a measurement of all the configurations that are invoked by the benchmarking tool. The measurements were conducted 5 to 20 times and the mean of all measurements for each configuration was recoded.

>>>>

\title{Learning Actionable Analytics 
 (with applications for reducing defects and reducing runtimes)}
\author{
% You can go ahead and credit any number of authors here,
% e.g. one 'row of three' or two rows (consisting of one row of three
% and a second row of one, two or three).
%
% The command \alignauthor (no curly braces needed) should
% precede each author name, affiliation/snail-mail address and
% e-mail address. Additionally, tag each line of
% affiliation/address with \affaddr, and tag the
% e-mail address with \email.
%
% 1st. author
Rahul Krishna, Tim Menzies, Xipeng Shen\\
       \affaddr{Computer Science, NcState, USA}\\
       \{i.m.ralk, tim.menzies, xipengshen\}@gmail.com
% 3rd. author
 % use '\and' if you need 'another row' of author names
% 4th. author
\and
 Andrian Marcus\\
       \affaddr{Computer Science, UtDallas, USA}\\
       {amarcus@utdallas.edu} }

  \pagestyle{plain}
\begin{document}
  \maketitle

  % make the title area

@@

\section{Introduction}
Business  users are now demanding that data mining tools
be augmented with tools to support  business-level
interpretation of that data. For example,
at a  panel on software analytics at ICSE'12,
industrial practitioners lamented the state of the art in data mining
and software engineering~\cite{menzies12a}. Panelists commented that
``prediction is all well and good, but what about decision
making?''. That is, these panelists are more interested in the interpretations
that follow the mining, rather than just  the mining. For example:
\bi
\item Instead of just accepting  {\em predictions} on how many 
 software defects
to expect,  business users might now demand a {\em plan} to
reduce the likelihood of those defects;
\item Instead of just accepting {\em predictions} on the runtime
time of their software, business users might now demand
a {\em plan} to reduce that runtime.
\ei
In these two points, we say that a ``plan'' is a  set of proposed changes in order to better achieve some goal. 
Generating {\em plans} is a  very different task the standard
data mining tasks of {\em predicting} some discrete or numeric class.
Predictors
and classifiers only tell us what {\em is} while planners 
  tell us {\em what to change}. For the kind of business user
discussed above, planners are the preferred to mere prediction systems
since, as one gruff user once shouted at us:
\begin{quote}
{\em ``Don't tell me what \underline{is};
tell me what to \underline{change}.''}
\end{quote}
In order to satisfy the demands of this new kind of business user, we  study the generation of {\em plans} in the context
of software analytics.  
This paper presents
HOW, a novel  {\em instance-based planner} that runs in near-linear time.
In the experiments shown below, we see that HOW's plans can
significantly reduce the expected
value of:
\bi
\item Software defects in certain object-oriented JAVA systems listed in \tion{oo}:  Ant Camel, Ivy, Jedit, JBeans, Log4, Lucene, Poi, Synapse, Velocity, Xalan, Xerces;
\item The runtime of the software systems discussed in \tion{cpu}: Berkeley DB, Apache, SQLite, LLVM, 
  x264.
\ei
HOW's instance-based reasoning  makes its conclusions by
studying the space of examples. This is a very different approach to
{\em model-based planners} that use the model
as sub-routine within a Monte Carlo simulation~\cite{me07f} or the evolutionary 
programming methods explored by the search-based software
engineering community~\cite{krall14,harman12dec}. 
We find   HOW's instance-based approach useful since it can be applied
to domains that lack a model, but possess a historical log of observed behaviours.

Instance-based reasoning is also useful if building and certifying a model   takes a long time. Such slow-to-build models exist in software engineering. For  example, in previous work~\cite{me07f} we used models
developed by Boehm's group at the University of Southern California.
Some of those models took decades to develop and mature (from 1981~\cite{boehm81} to 2000~\cite{boehm00b}).  This is not an isolated example. One of us (Menzies) teaches a graduate class in automated model-based software engineering. 
Students in that class are asked to find (or build) a non-trivial model of
some software process (where ``non-trivial'' means something that processes
a  model at least an order of magnitude more complex than Brooks' Law). 
While the  number of such models has been growing in recent years\footnote{See the
catalogue at goo.gl/nv2AVK.}, 
the total number remains very small-- which is an observation consistent with
SE models typically require  much effort to build and certify.

Finally, even when there is an existing model, instance-based reasoning can be useful.
Sometimes  it is easier to access instances of a model's behaviour than the model
itself. For example, in prior work with Martin  Feather from the Jet Propulsion 
Laboratory~\cite{fea02a},  our research partner could not share a
propriety model from within the NASA firewalls. However, they could share 
logs of the input/output of that model.

In summary, this paper makes five contributions:
\be
\item A general framework for augmenting a prediction system such that
it becomes a planning system;
\item A  test that recognizes  when we should {\em not}  plan: in summary, our kind of planning is
not recommended in domains that  generate poorly performing predictors;
\item A general verification rig for assessing the value of plans generated within this rig.
\item The HOW  instance-based planner.
\item A proof that HOW runs in near-linear time.
\item An experimental validation that the HOW planner is useful for reducing the
expected values of software defects and software runtimes (at least, for the data sets studied here).
\ee
The rest of this paepr is structued as follows.

\section{From ``Prediction'' to ``Planning''}
A major  problem with building a planning mechanism in software engineering is ``how to verify the generated
plans?''. For example,  in the experiments shown below,  HOW comments  on how to reduce
software defects by adjusting certain static code parameters of the code\footnote{Such parameters can be automatically adjusted using code refactoring
tools, as shown by Binkely et al. in 2006~\cite{Binkley2006} and more recently by Mkaouer et al. in 2015~\cite{Mkaouer15}.}. How are we to verify that HOW's recommendations actually work?

Some organizations have the resources to 
run repeated trials to assess the effect of different treatments.
For example, in one remarkable recent study, Bente et al. report results
were the same specification was developed into a product by four different organizations~\cite{Anda2009}. Given those kind of resources, it would be possible
to (say) take a code base, assign it to $N$ different teams, apply
HOW's plans the code used by  $N/2$ teams, then check in 12 months time
if those halves of the teams have fewer defects than the other.

Very few industrial or research groups have access
to the kinds of resources using by Bente et al. (evidence: in the size years since the
publication of that work, we know of only one of two similar studies). Hence, it
we wish to verify HOW's plans, then some other approach is required that is less
resource intensive.

We propose using prediction systems (built by data mining) as a verification
tool for planners. In domains with data, from which data miners can build
predictors of adequate performance, then those predictors can assess the value
of the changes proposed by HOW.
As shown in \fig{work}, 
project data is divided into two disjoint sets {\em train} and {\em test}
(so {\em train} $\cup$ {\em test} $=\;\emptyset$).
Next, from the train set, we build both a {\em planner} as well
as a {\em predictor}. Our general framework does not   commit to any particular  choice
of {\em planner} or {\em predictor}. For the purposes of this discussion 
our {\em planner}
will be HOW (discussed below)and our
our {\em predictors} will be the Random Forest and Random Forest
Regressor\footnote{Decision tree learners find the attribute whose value's
select for for sections of the data with least variation in the class
variable. Decision tree learners then recurse on each selection~\cite{breiman84}. A random forest is a set of decision trees, each built
from some random subset of the rows and columns. The conclusions
of the forest are a ``vote'' generated by all the trees~\cite{Breiman2001}. 
A standard Random Forest predicts for a discrete class variable while
a Random Forest Regressor generates numeric predictions.} taken from the SciKit
Learn toolkit~\cite{Pedregosa2012}.

Turning now to the {\em test} data, this is based to the {\em predictor}
to generate a baseline {\em before} performance score.
Note that if our {\em predictors} fail to perform adequately on the test data,
then we cannot trust them to comment on our plans. Accordingly,
if that performance is unsatisfactory, we abort.
Else, we (1)~apply the {\em planner} to alter the {\em test} data;
then (2)~apply the {\em predictor} to the altered data {\em test'};
then (3)~return the performance delta $\Delta$.

The clear advantage of this approach is that it can be readily and automatically
applied to domains with historical data. That said, the drawback with this approach
is that our assessment of the plans is only as good as the predictor. To 
mitigate for this {\em problem of poor planners}, it is wise to demand high standards form the {\em
satisfactory} threshold used in this \fig{frame}.

\begin{figure}[!t]
\small
\[  
\begin{array}{r}
project\\
data
\end{array} 
\left\{\begin{array}{l}\mathit{train}
        \left\{\begin{array}{l}
                \mathrm{learn\;a\;}\mathrm{predictor\;}\mathrm{e.g.\;Random\;Forest}\\
                \mathrm{learn\;a\;}\mathrm{planner\;}\mathrm{e.g.\; HOW}
              \end{array}\right.
       \\
      ~\\
\mathit{test}  
    \left\{\begin{array}{l@{~}l}
           \mathit{before}& =\mathrm{predictor}(\mathit{test})\\
           \mathrm{\bf if\;}\mathit{before} & <  \mathit{satisfactory}\\
           \mathrm{\bf then}  & \mathrm{{\bf return}\; 0}\\~\\
           \mathrm{\bf else} &
           \left\{
            \begin{array}{l}
                \mathit{test'} = \mathrm{planner}(\mathit{test})\\
                \mathit{after} =\mathrm{predictor}(\mathit{test'})\\ 
                \mathrm{{\bf return}\;} \mathit{after} /  \mathit{before}
            \end{array}
          \right.
   \end{array}\right.
\end{array} \right. 
\]

\caption{How to use {\em predictors} for {\em planning}.}\label{fig:work}
\end{figure}

This study explores data sets from two types of domains:
\bi
\item
In one type, we 
build plans to change static code attributes in order to reduce defects in
Java classes.  Here,,
performance can be measured in terms of recall (percent of faulty classes in
the test data detected
by the {\em predictor}). Using engineering judgement, we define satisfactory recall as anything less than 70\%. 
\item 
In the other type, we   build
plans to change configurations settings within a Makefile such that the resulting compiled
code will run faster. Here, performance can be measured in terms of difference
between the predicted runtime $p$ of test case items and their actual runtimes $a$
using  $s= 1- \frac{abs(a - p)}{a}$.
Using engineering judgement, we define satisfactory $s$ as anything less than 70\%. 
\ei
This paper explores 12 static code data sets
Ant, Camel, Ivy, Jedit, JBeans, Log4, Lucene, Poi, Synapse, Velocity, Xalan, Xerces\footnote{Available from the object-oriented defects section of the PROMISE respository openscience.us/repo/defect/ck.} and five configuration data sets Berkeley DB, Apache, SQLite, LLVM,
  x264\footnote{Available from the performance prediction section of PROMISE
  http://openscience.us/repo/performance-predict.}. For each of these data sets,
  we generated 25 {\em before} scores from
  random split train:test splits in the ratio 4:1 train.
 The median {\em before} recalls from  the static code data sets was:
%issue24
\[\mathrm{recall}=\{98,88,86,84,82,81,79,78,78,\overbrace{55,23,20}^{\mathrm{JBeans,Poi,Xalan}}\}\]
The median {\em before} $e$  scores from  the configuration data was
%issue 24
\[s=\{98,97,95,88,92\}\]
Note that we need to abort planning in only $3/17$ of our data sets.

This paper also explores five configuration data sets Berkeley DB, Apache, SQLite, LLVM,
  x264. The median {\em before} scores from these data sets (generated in 
-- which means we cannot generate plans for $3/12$ of our data sets. However, as shown
below, we were able to generate useful plans for the other $9/12$ of the data.

In the other type,

then studied to generate a prediction
on the test data. If the performance on that test data is not
satisfactory, thempredictions:
a {\em before} prediction (on the raw test data); and  an {\em after}
prediction (that is  applying the policies learned from the training data).

The improvement delta $\Delta$ is then computed by dividing {\em after} and {\em before}

This paper presents a novel policy generator called ``HOW''.
HOW is a near-linear time  instance-based
local search algorithm that looks for improvement trends in the data,
then adjusting examples to nudge them along that trend.
The experiments of this paper apply HOW to learning policies for 
software defect reduction and for compiling software that runs faster.
At least for the data explored here, HOW's reductions in defects and runtimes
are very large-- up to 50\% fewer defects and up to 50\% faster runtimes.

@@

\subsection{Defect Data Set}
The data was obtained from the PROMISE repository\footnote{Promise Repository: \url{openscience.us/repo}}. For the defect data we investigated 32 releases from 11 open source Java projects defined by the metrics highlighted in figure ~\ref{fig:ck}: \kw{Apache Ant} (1.5 -- 1.7), \kw{Apache Camel} (1.2 -- 1.6), \kw{Apache Ivy} (1.1 -- 2.0), \kw{JEdit} (4.1 -- 4.3), \kw{Apache Log4j} (1.0 -- 1.2), \kw{Apache Lucene} (2.0 -- 2.2), \kw{PBeans} (1.0 and 2.0), \kw{Apache POI} (2.0 -- 3.0), \kw{Apache Synapse} (1.0 -- 1.2), \kw{Apache Velocity} (1.4 -- 1.6), and \kw{Apache Xalan-Java} (2.5 -- 2.7).

@@

{\bfseries RQ3} asks if the aforementioned method can be expanded to other kinds of data. To answer this, we have used the performance prediction data set collected by Siegmund et al. for their ICSE '12 paper~\cite{sven12}. The data sets have been deployed as a part of the \textsc{SPLConqueror} tool\footnote{\url{fosd.de/SPLConqueror}}. This data set contains six examples of real world configurable systems that cover a broad spectrum of scenarios, see figure \ref{fig:cpm} for a summary. The systems have different characteristics (lines of code: 40 thousand to 300 thousand; Valid configurations: 192 to about 3.9 million), different implementation languages (C, C++, and Java), different mechanism (conditional compilation, configuration files, and command-line options), and different domains (Web servers, compilers, encoders, and databases). The data contains a measurement of all the configurations that are invoked by the benchmarking tool. The measurements were conducted 5 to 20 times and the mean of all measurements for each configuration was recoded.
